---
title: "DDOS Project EDA"
author: "Firass Elhouat"
date: "2025-03-14"
output: pdf_document
---

## 1. Introduction

In today’s digital landscape, Distributed Denial-of-Service (DDoS) attacks have emerged as one of the most prevalent and disruptive cybersecurity threats facing business and organizations of all sizes. This includes government agencies, educational institutions, healthcare, and any business that relies on servers and network access.

These attacks are designed to overwhelm a target’s/victim’s servers or network infrastructure with a flood of illegitimate traffic, rendering online services inaccessible to legitimate users and other stakeholders. The costs incurred during these attacks tend to vary depending on the duration and attackers motives. For instance, an insight report published in 2023 by Zayn Group, found that DDoS attacks on average last around 68 minutes, with unprotected organizations having to payout around \$6,000 per minute of each attack, [(Zayo Group (2024))](https://www.zayo.com/newsroom/average-ddos-attack-cost-businesses-nearly-half-a-million-dollars-in-2023-according-to-new-zayo-data/).

This totals to around \$408,000 per attack, that a business loses, which takes into account lost revenue, cost of detection, recovery, and cybersecurity preventives. While the immediate impact of a DDoS attack manifest as downtime or service disruptions, the broader consequences extend far beyond revenue loss. Businesses suffer from brand reputation damage, operational disruptions, legal ramifications, customer churn, and costly mitigation efforts.

A notable example is Sony’s PlayStation Network (PSN), which has been a frequent target of DDoS and other Malware attacks, most infamously in 2011, when a large-scale attack brought down the network for nearly a month, [(Garcia, D.M. (2021))](https://wsswired.com/4837/entertainment-3/the-2011-playstation-network-hack-what-actually-happened/) . This attack not only resulted in financial losses due to service interruptions and compensations to users but also severely impacted customer trust and brand loyalty. Many frustrated gamers switched to Steam (PC Gaming Platform) and Microsoft’s Xbox Live, which was perceived as more stable and secure at the time.

The PlayStation Network attack highlighted the long-term effects of DDoS attack, where the loss of consumer confidence led to a direct increase in competitors market share. Sony had to invest heavily in cybersecurity upgrades, public relations efforts, and damage control to regain trust, but the attack had already left a lasting impression on its user baser.

To better understand and model DDoS attacks, we will be utilizing a publicly available dataset from [Kaggle](https://www.kaggle.com/datasets/aymenabb/ddos-evaluation-dataset-cic-ddos2019/data){.uri}. This dataset contains extensive network traffic data, including packet details, communication protocols, and attack indicators. By applying Exploratory Data Analysis (EDA), we aim to identify patterns that distinguish normal traffic from malicious DDoS activity. Through this analysis, we aim to develop a predictive model that can detect potential DDoS threats in real time, helping us fully understand how these attacks occur, and what are some possible mitigation efforts that can be implemented to reduce the significant effects, and how each variable in our dataset explains our target variable.

## **2. Dataset**

To better understand and model DDoS attacks, we will use the CIC-DDoS2019 dataset that was generated using a [CICFlowMeter-V3](https://www.unb.ca/cic/research/applications.html#CICFlowMeter) which contains 225,745 rows, and 85 various variables that help differentiate between normal and attack-related traffic. The dataset includes these defined attributes/variables:

-   **Identifier and date variables:** These are most likely to be dropped as they are not useful during the modeling phase of our project.

    -   **flow_id**

    -   **source_ip**

    -   **source_port**

    -   **destination_ip**

    -   **destination_port**

    -   **Time stamp**

-   **Categorical Variables:**

    -   **protocol: Denotes the protocol used,** with three levels; **0, 6 and 17**

-   **Continuous Variables:**

    -   **flow_duration:** The duration of the flow in microseconds.

    -   **total_fwd_packets:** The total packets in the forward direction.

    -   **total_backward_packets:** The total packets in the backward direction.

    -   **total_length_of_fwd_packets:** The total size of a packet in the forward direction.

    -   **total_length_of_bwd_packets:** The total size of a packet in the backward direction.

    -   **Forward (FWD) Packet Length Metrics:**

        -   **fwd_packet_length_max:** The maximum size of a packet in the forward direction.

        -   **fwd_packet_length_min:** The minimum size of a packet in the forward direction.

        -   **fwd_packet_length_mean:** The mean size of a packet in the forward direction.

        -   **fwd_packet_length_std:** The Standard deviation size of a packet in the forward direction.

    -   **Backward (BWD) Packet Length Metrics:**

        -   **bwd_packet_length_max:** The maximum size of a packet in the backward direction.

        -   **bwd_packet_length_min:** The minimum size of a packet in the backward direction.

        -   **bwd_packet_length_mean:** The mean size of a packet in the backward direction.

        -   **bwd_packet_length_std:** The standard deviation size of a packet in the backward direction.

    -   **flow_bytes_s:** The number of flow bytes per second.

    -   **flow_packets_s:** The number of flow packets per second.

    -   **Flow Inter-Arrival Time (IAT) Metrics:**

        -   **flow_iat_mean:** The average inter-arrival time between packets within a flow.

        -   **flow_iat_std:** The standard deviation inter-arrival time between packets within a flow.

        -   **flow_iat_max:** The maximum inter-arrival time between packets within a flow.

        -   **flow_iat_min:** The minimum inter-arrival time between packets within a flow.

    -   **Forward (FWD) Inter-Arrival Time (IAT) Metrics:**

        -   **fwd_iat_total:** The total inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_mean:** The average inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_std:** The standard deviation inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_max:** The maximum inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_min:** The minimum inter-arrival time between packets sent in the forward direction.

    -   **Backward (BWD) Inter-Arrival Time (IAT) Metrics:**

        -   **bwd_iat_total:** The total inter-arrival time between packets sent in the forward direction.

        -   **bwd_iat_mean:** The average inter-arrival time between packets sent in the forward direction.

        -   **bwd_iat_std:** The standard deviation inter-arrival time between packets sent in the forward. direction.

        -   **bwd_iat_max:** The maximum inter-arrival time between packets sent in the forward direction.

        -   **bwd_iat_min:** The minimum inter-arrival time between packets sent in the forward direction.

    -   **fwd_psh_flags:** Number of time the PSH (Push) flag was set in packets travelling in the forward direction.

    -   **bwd_psh_flags:** Number of time the PSH (Push) flag was set in packets travelling in the backward direction.

    -   **fwd_urg_flags:** Number of time the URG (Urgent) flag was set in packets travelling in the forward direction.

    -   bwd_urg_flags: Number of time the URG (Urgent) flag was set in packets travelling in the forward direction.

    -   **fwd_header_length_41:** The total bytes used for headers in the forward direction.

    -   **bwd_header_length:** The total bytes used for headers in the backward direction.

    -   **fwd_packets_s:** Number of forward packets transmitted per second.

    -   **bwd_packets_s:** Number of backward packets transmitted per second.

    -   **Packet Length Metrics:**

        -   **min_packet_length:** The minimum length of a packet.

        -   **max_packet_length:** The maximum length of a packet.

        -   **packet_length_mean:** The average length of a packet.

        -   **packet_length_std:** The standard deviation length of a packet.

        -   **packet_length_variance:** The variance length of a packet.

    -   **fin_flag_count:** The number of packets with FIN (Finish) flag, which signals the termination of a connection.

    -   **syn_flag_count:** The number of packets with SYN (Synchronize) flag, used to initiate a TCP connection.

    -   **rst_flag_count:** The number of packets with RST (Reset) flag, used to abruptly terminate a connection.

    -   **psh_flag_count:** The number of packets with PSH (Push) flag, which tells the receiver to process data immediately rather than buffering it.

    -   **ack_flag_count:** The number of packets with ACK (Acknowledgment) flag, which confirms the receipt data.

    -   **urg_flag_count:** The number of packets with URG (Urgent) flag, indicating that certain data should be processed immediately.

    -   **cwe_flag_count:** Will be renamed to correct name : CWR (Congestion window reduced) flag, used in TCP congestion control to signal a reduced congestion window.

    -   **ece_flag_count:** The number of packets with ECE (Explicit Congestion Notification Eco) flag, indicates network congestion.

    -   **down_up_ratio:** Download and upload ratio.

    -   **average_packet_size:** The average size of a packet.

    -   **avg_fwd_segment_size:** The average size of data segments in the forward direction.

    -   **avg_bwd_segment_size:** The average size of data segments in the backward direction.

    -   **fwd_header_length_62:** Length of a packet header in the forward direction.

    -   **fwd_avg_bytes_bulk:** Average number of bulk bytes rate in the forward direction.

    -   **fwd_avg_packets_bulk:** Average number of bulk packets rate in the forward direction.

    -   **fwd_avg_bulk_rate:** Average number of bulk rate in the forward direction.

    -   **bwd_avg_bytes_bulk:** Average number of bulk bytes rate in the backward direction.

    -   **bwd_avg_packets_bulk:** Average number of bulk packets rate in the backward direction.

    -   **bwd_avg_bulk_rate:** Average number of bulk rate in the backward direction.

    -   **subflow_fwd_packets:** The average number of packets in a sub flow in the forward direction.

    -   **subflow_fwd_bytes:** The average number of bytes in a sub flow in the forward direction.

    -   **subflow_bwd_packets:** The average number of packets in a sub flow in the backward direction.

    -   **subflow_bwd_bytes:** The average number of bytes in a sub flow in the backward direction.

    -   **init_win_bytes_forward:** The total number of bytes sent in initial window in the forward direction.

    -   **init_win_bytes_backward:** The total number of bytes sent in initial window in the **backward** direction.

    -   **act_data_pkt_fwd:** Actual number of packets sent in the forward direction.

    -   **min_seg_size_forward:** Minimum size of the data segment sent in the forward direction.

    -   **Active Metrics :**

        -   **active_mean:** The average time a flow was active before becoming idle.

        -   **active_std:** The standard deviation time a flow was active before becoming idle.

        -   **active_max:** The maximum time a flow was active before becoming idle.

        -   **active_min:** The minimum time a flow was active before becoming idle.

    -   

    ```{=html}
    <!-- -->
    ```
    -   **IDLE Metrics:**

        -   **idle_mean:** The average time a flow was idle before becoming active.

        -   **idle_std:** The standard deviation time a flow was idle before becoming active.

        -   **idle_max:** The maximum time a flow was idle before becoming active.

        -   **idle_min:** The minimum time a flow was idle before becoming active.

    -   

-   **Response Variable:**

    -    **label:**

        -   **BENIGN:** Normal legitimate network traffic

        -   **DDoS:** Malicious network traffic.

```{r setup, include=FALSE}
# --- loading libraries 
library(tidyverse) # for extra functions 
library(janitor) # for column name cleaning 
library(viridis) # for color palettes 
library(corrplot) # for correlation plot 
library(ggnewscale) # for plots 

```

## 

```{r}
# ---- loading data 
DDoS_dataset <- read_csv("Data/Friday-WorkingHours-Afternoon-DDos.pcap_ISCX.csv") %>% clean_names()

# view data 
print(DDoS_dataset)
```

```{r}
# unique Values
length(unique(DDoS_dataset$label))
length(unique(DDoS_dataset$fwd_header_length))

# remove unique ID columns 
# (flow_id, source_ip, source_port, destination_ip, destination_port)
```

## 3. Data Preprocessing 

Remove unique ID columns

-   flow_id

-   source_ip

-   source_port

-   destination_ip

-   destination_port

-   timestamp

```{r}
# --- remove unwanted variables 
DDoS_dataset <- DDoS_dataset %>%
  select(-flow_id, -source_ip, 
         -source_port, -destination_ip, 
         -destination_port, -timestamp)
```

Checking and removing null rows

```{r}
# --- check for null values 
sum(is.na(DDoS_dataset))
```

## 4. Data Analysis 

### 4.1 Correlation Matrix 

In this section, creating a correlation matrix

```{r}
# 
DDoS_dataset <- DDoS_dataset %>%
  mutate(label = recode(label, "DDoS" = 1, "BENIGN" = 0))

# assigning columns as numeric 
DDoS_dataset <- DDoS_dataset %>%
  mutate(across(-c(1, 79), as.numeric))

# Select the columns: 1 to 40 and 79
DDoS_dataset %>%
  select(1:40, 79) %>%  # Select columns 1 to 39 and column 79
  select(which(sapply(., sd) != 0)) %>%
  cor() %>% corrplot(method = "color", 
        tl.cex = 0.5, 
         number.cex = 0.5,
         tl.srt = 45,
          tl.col = "black",
         order = "hclust",
         addCoef.col = "black",  # Display correlation values
           number.digits = 1)
```

```{r}
# Select the columns: 41 to 79
DDoS_dataset %>%
  select(41:79) %>%  # Select columns 1 to 39 and column 79
  select(which(sapply(., sd) != 0)) %>%
  cor() %>% corrplot(method = "color", 
         tl.cex = 0.5, 
         number.cex = 0.5,
         tl.srt = 45,
          tl.col = "black",
         order = "hclust",
         addCoef.col = "black",  # Display correlation values
           number.digits = 1)

```




### 4.2 Data Viz 

```{r}
# Plot the distribution of the 'Target' variable

# bar plot of target distribution  
ggplot(DDoS_dataset, aes(x = as.factor(label))) +
  geom_bar(aes(fill = as.factor(label)), color = "black", alpha = 0.8, linewidth = 0.3) +
  labs(title = "Distribution of DDoS Attack Labels", x = "", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), 
                    labels = c("0" = "BENIGN", "1" = "DDoS"),
                    guide = guide_legend(title = NULL)) +  # Remove legend title
  scale_y_continuous(labels = scales::comma)


```


```{r}
table(DDoS_dataset$label, DDoS_dataset$protocol)


# bar plot of target and protocol distribution
ggplot(DDoS_dataset, aes(x = as.factor(protocol), fill = as.factor(label))) +
  geom_bar(color = "black", alpha = 0.8, linewidth = 0.3) +
  labs(title = "Distribution of DDoS Attack Labels by Protocol", x = "Protocol", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), 
                    labels = c("0" = "BENIGN", "1" = "DDoS"),
                    guide = guide_legend(title = NULL)) +
  scale_y_continuous(labels = scales::comma)

# Issues protocol variable 
# When "0", BENIGN is 54, while DDoS is 0. 
# When "6", BENIGN is 64793, while DDoS is 128027. 
# When "17", BENIGN is 32871, while DDoS is 0. 

```

```{r}
ggplot() + 
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 0, ], 
             aes(x = flow_duration, y = fwd_iat_total, color = "BENIGN"), 
             alpha = 0.6, size = 2) +  # Color for label 0
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 1, ], 
             aes(x = flow_duration, y = fwd_iat_total, color = "DDoS"), 
             alpha = 0.8, size = 2.5) +  # Color for label 1
  scale_color_manual(values = c("BENIGN" = "blue", "DDoS" = "red")) +  # Custom color mapping
  theme_minimal() +
  labs(title = "Flow Duration vs FWD IAT Total", x = "Flow Duration", y = "FWD IAT Total") + 
  theme(legend.title = element_blank()) 

```



```{r}

ggplot() + 
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 0, ], 
             aes(x = total_fwd_packets, y = total_backward_packets, color = "BENIGN"), 
             alpha = 0.6, size = 2) +  # Color for label 0
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 1, ], 
             aes(x = total_fwd_packets, y = total_backward_packets, color = "DDoS"), 
             alpha = 0.8, size = 2.5) +  # Color for label 1
  scale_color_manual(values = c("BENIGN" = "blue", "DDoS" = "red")) +  # Custom color mapping
  theme_minimal() +
  labs(title = "Total Forward Packets vs Total Backward Packets", 
       x = "Total Forward Packets", y = "Total Backward Packets") + 
  theme(legend.title = element_blank())  # Remove legend title

```

```{r}
# Plotting with ggplot
DDoS_dataset %>%
  select(contains("flag_count"), label) %>%
  group_by(label) %>%
  summarise(across(everything(), sum, na.rm = TRUE)) %>%
  pivot_longer(cols = -label, 
               names_to = "flag_type", 
               values_to = "flag_count") %>%
  ggplot(aes(x = flag_type, y = flag_count, fill = as.factor(label))) + 
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) + 
  scale_fill_manual(values = c("red", "blue")) + 
  theme_minimal() +
  labs(title = "Aggregated Flag Counts by Label", 
       x = "Flag Type", 
       y = "Count of Flags", 
       fill = "Label") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```


```{r}

ggplot(DDoS_dataset, aes(x = packet_length_std, y = packet_length_variance, colour = as.factor(label))) + 
  geom_point(alpha = 0.6, size = 2, position = "identity") +  # Set transparency for all points
  scale_color_manual(values = c("1" = "red", "0" = "blue")) +  # Reverse color mapping
  theme_minimal() +  # Clean theme
  labs(title = "Packet Length Std vs Packet Length Variance", x = "Packet Length Std", y = "Packet Length Variance Total") +  # Labels
  theme(legend.title = element_blank(),  # Remove legend title
        legend.key = element_rect(color = "black"))
```

## 5. Reference list 

-   Zayo Group (2024) *Average ddos attack cost businesses nearly half a million dollars in 2023, according to New Zayo Data: Press release: Zayo*, *Zayo.com*. Available at: <https://www.zayo.com/newsroom/average-ddos-attack-cost-businesses-nearly-half-a-million-dollars-in-2023-according-to-new-zayo-data/> (Accessed: 18 March 2025).

-   Garcia, D.M. (2021) *The 2011 PlayStation Network Hack – what actually happened?*, *WestSide Story*. Available at: <https://wsswired.com/4837/entertainment-3/the-2011-playstation-network-hack-what-actually-happened/> (Accessed: 18 March 2025).
