---
title: "DDOS Project EDA"
author: "Firass Elhouat"
date: "2025-03-14"
output: pdf_document
---

## 1. Introduction

In today’s digital landscape, Distributed Denial-of-Service (DDoS) attacks have emerged as one of the most prevalent and disruptive cybersecurity threats facing business and organizations of all sizes. This includes government agencies, educational institutions, healthcare, and any business that relies on servers and network access.

These attacks are designed to overwhelm a target’s/victim’s servers or network infrastructure with a flood of illegitimate traffic, rendering online services inaccessible to legitimate users and other stakeholders. The costs incurred during these attacks tend to vary depending on the duration and attackers motives. For instance, an insight report published in 2023 by Zayn Group, found that DDoS attacks on average last around 68 minutes, with unprotected organizations having to payout around \$6,000 per minute of each attack, [(Zayo Group (2024))](https://www.zayo.com/newsroom/average-ddos-attack-cost-businesses-nearly-half-a-million-dollars-in-2023-according-to-new-zayo-data/).

This totals to around \$408,000 per attack, that a business loses, which takes into account lost revenue, cost of detection, recovery, and cybersecurity preventives. While the immediate impact of a DDoS attack manifest as downtime or service disruptions, the broader consequences extend far beyond revenue loss. Businesses suffer from brand reputation damage, operational disruptions, legal ramifications, customer churn, and costly mitigation efforts.

A notable example is Sony’s PlayStation Network (PSN), which has been a frequent target of DDoS and other Malware attacks, most infamously in 2011, when a large-scale attack brought down the network for nearly a month, [(Garcia, D.M. (2021))](https://wsswired.com/4837/entertainment-3/the-2011-playstation-network-hack-what-actually-happened/) . This attack not only resulted in financial losses due to service interruptions and compensations to users but also severely impacted customer trust and brand loyalty. Many frustrated gamers switched to Steam (PC Gaming Platform) and Microsoft’s Xbox Live, which was perceived as more stable and secure at the time.

The PlayStation Network attack highlighted the long-term effects of DDoS attack, where the loss of consumer confidence led to a direct increase in competitors market share. Sony had to invest heavily in cybersecurity upgrades, public relations efforts, and damage control to regain trust, but the attack had already left a lasting impression on its user baser.

To better understand and model DDoS attacks, we will be utilizing a publicly available dataset from [Kaggle](https://www.kaggle.com/datasets/aymenabb/ddos-evaluation-dataset-cic-ddos2019/data){.uri}. This dataset contains extensive network traffic data, including packet details, communication protocols, and attack indicators. By applying Exploratory Data Analysis (EDA), we aim to identify patterns that distinguish normal traffic from malicious DDoS activity. Through this analysis, we aim to develop a predictive model that can detect potential DDoS threats in real time, helping us fully understand how these attacks occur, and what are some possible mitigation efforts that can be implemented to reduce the significant effects, and how each variable in our dataset explains our target variable.

## **2. Dataset**

To better understand and model DDoS attacks, we will use the CIC-DDoS2019 dataset that was generated using a [CICFlowMeter-V3](https://www.unb.ca/cic/research/applications.html#CICFlowMeter) which contains 225,745 rows, and 85 various variables that help differentiate between normal and attack-related traffic. The dataset includes these defined attributes/variables:

-   **Identifier and date variables:** These are most likely to be dropped as they are not useful during the modeling phase of our project.

    -   **flow_id**

    -   **source_ip**

    -   **source_port**

    -   **destination_ip**

    -   **destination_port**

    -   **Time stamp**

-   **Categorical Variables:**

    -   **protocol: Denotes the protocol used,** with three levels; **0, 6 and 17**
    -   **fin_flag_count:** The number of packets with FIN (Finish) flag, which signals the termination of a connection, with two levels, 1 and 0.
    -   **syn_flag_count:** The number of packets with SYN (Synchronize) flag, used to initiate a TCP connection, with two levels, 1 and 0.
    -   **rst_flag_count:** The number of packets with RST (Reset) flag, used to abruptly terminate a connection, with two levels, 1 and 0.
    -   **psh_flag_count:** The number of packets with PSH (Push) flag, which tells the receiver to process data immediately rather than buffering it, with two levels, 1 and 0.
    -   **ack_flag_count:** The number of packets with ACK (Acknowledgment) flag, which confirms the receipt data, with two levels, 1 and 0.
    -   **urg_flag_count:** The number of packets with URG (Urgent) flag, indicating that certain data should be processed immediately, with two levels, 1 and 0.
    -   **cwe_flag_count:** Will be renamed to correct name : CWR (Congestion window reduced) flag, used in TCP congestion control to signal a reduced congestion window, with two levels, 1 and 0.
    -   **ece_flag_count:** The number of packets with ECE (Explicit Congestion Notification Eco) flag, indicates network congestion, with two levels, 1 and 0.
    -   **fwd_psh_flags:** Number of time the PSH (Push) flag was set in packets travelling in the forward direction, with two levels, 1 and 0.
    -   **bwd_psh_flags:** Number of time the PSH (Push) flag was set in packets travelling in the backward direction, with two levels, 1 and 0.
    -   **fwd_urg_flags:** Number of time the URG (Urgent) flag was set in packets travelling in the forward direction, with two levels, 1 and 0.
    -   bwd_urg_flags: Number of time the URG (Urgent) flag was set in packets travelling in the forward direction, with two levels, 1 and 0.

-   **Continuous Variables:**

    -   **flow_duration:** The duration of the flow in microseconds.

    -   **total_fwd_packets:** The total packets in the forward direction.

    -   **total_backward_packets:** The total packets in the backward direction.

    -   **total_length_of_fwd_packets:** The total size of a packet in the forward direction.

    -   **total_length_of_bwd_packets:** The total size of a packet in the backward direction.

    -   **Forward (FWD) Packet Length Metrics:**

        -   **fwd_packet_length_max:** The maximum size of a packet in the forward direction.

        -   **fwd_packet_length_min:** The minimum size of a packet in the forward direction.

        -   **fwd_packet_length_mean:** The mean size of a packet in the forward direction.

        -   **fwd_packet_length_std:** The Standard deviation size of a packet in the forward direction.

    -   **Backward (BWD) Packet Length Metrics:**

        -   **bwd_packet_length_max:** The maximum size of a packet in the backward direction.

        -   **bwd_packet_length_min:** The minimum size of a packet in the backward direction.

        -   **bwd_packet_length_mean:** The mean size of a packet in the backward direction.

        -   **bwd_packet_length_std:** The standard deviation size of a packet in the backward direction.

    -   **flow_bytes_s:** The number of flow bytes per second.

    -   **flow_packets_s:** The number of flow packets per second.

    -   **Flow Inter-Arrival Time (IAT) Metrics:**

        -   **flow_iat_mean:** The average inter-arrival time between packets within a flow.

        -   **flow_iat_std:** The standard deviation inter-arrival time between packets within a flow.

        -   **flow_iat_max:** The maximum inter-arrival time between packets within a flow.

        -   **flow_iat_min:** The minimum inter-arrival time between packets within a flow.

    -   **Forward (FWD) Inter-Arrival Time (IAT) Metrics:**

        -   **fwd_iat_total:** The total inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_mean:** The average inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_std:** The standard deviation inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_max:** The maximum inter-arrival time between packets sent in the forward direction.

        -   **fwd_iat_min:** The minimum inter-arrival time between packets sent in the forward direction.

    -   **Backward (BWD) Inter-Arrival Time (IAT) Metrics:**

        -   **bwd_iat_total:** The total inter-arrival time between packets sent in the forward direction.

        -   **bwd_iat_mean:** The average inter-arrival time between packets sent in the forward direction.

        -   **bwd_iat_std:** The standard deviation inter-arrival time between packets sent in the forward. direction.

        -   **bwd_iat_max:** The maximum inter-arrival time between packets sent in the forward direction.

        -   **bwd_iat_min:** The minimum inter-arrival time between packets sent in the forward direction.

    -   **fwd_header_length_41:** The total bytes used for headers in the forward direction.

    -   **bwd_header_length:** The total bytes used for headers in the backward direction.

    -   **fwd_packets_s:** Number of forward packets transmitted per second.

    -   **bwd_packets_s:** Number of backward packets transmitted per second.

    -   **Packet Length Metrics:**

        -   **min_packet_length:** The minimum length of a packet.

        -   **max_packet_length:** The maximum length of a packet.

        -   **packet_length_mean:** The average length of a packet.

        -   **packet_length_std:** The standard deviation length of a packet.

        -   **packet_length_variance:** The variance length of a packet.

    -   **down_up_ratio:** Download and upload ratio.

    -   **average_packet_size:** The average size of a packet.

    -   **avg_fwd_segment_size:** The average size of data segments in the forward direction.

    -   **avg_bwd_segment_size:** The average size of data segments in the backward direction.

    -   **fwd_header_length_62:** Length of a packet header in the forward direction.

    -   **fwd_avg_bytes_bulk:** Average number of bulk bytes rate in the forward direction.

    -   **fwd_avg_packets_bulk:** Average number of bulk packets rate in the forward direction.

    -   **fwd_avg_bulk_rate:** Average number of bulk rate in the forward direction.

    -   **bwd_avg_bytes_bulk:** Average number of bulk bytes rate in the backward direction.

    -   **bwd_avg_packets_bulk:** Average number of bulk packets rate in the backward direction.

    -   **bwd_avg_bulk_rate:** Average number of bulk rate in the backward direction.

    -   **subflow_fwd_packets:** The average number of packets in a sub flow in the forward direction.

    -   **subflow_fwd_bytes:** The average number of bytes in a sub flow in the forward direction.

    -   **subflow_bwd_packets:** The average number of packets in a sub flow in the backward direction.

    -   **subflow_bwd_bytes:** The average number of bytes in a sub flow in the backward direction.

    -   **init_win_bytes_forward:** The total number of bytes sent in initial window in the forward direction.

    -   **init_win_bytes_backward:** The total number of bytes sent in initial window in the **backward** direction.

    -   **act_data_pkt_fwd:** Actual number of packets sent in the forward direction.

    -   **min_seg_size_forward:** Minimum size of the data segment sent in the forward direction.

    -   **Active Metrics :**

        -   **active_mean:** The average time a flow was active before becoming idle.

        -   **active_std:** The standard deviation time a flow was active before becoming idle.

        -   **active_max:** The maximum time a flow was active before becoming idle.

        -   **active_min:** The minimum time a flow was active before becoming idle.

    -   

    <!-- -->

    -   **IDLE Metrics:**

        -   **idle_mean:** The average time a flow was idle before becoming active.

        -   **idle_std:** The standard deviation time a flow was idle before becoming active.

        -   **idle_max:** The maximum time a flow was idle before becoming active.

        -   **idle_min:** The minimum time a flow was idle before becoming active.

    -   

-   **Response Variable:**

    -   **label:**

    ```         
    -   **BENIGN:** Normal legitimate network traffic

    -   **DDoS:** Malicious network traffic.
    ```

```{r setup, include=FALSE}
# --- loading libraries 
library(tidyverse) # for extra functions 
library(janitor) # for column name cleaning 
library(viridis) # for color palettes 
library(corrplot) # for correlation plot 
library(ggnewscale) # for extra plot functions 
library(car) # for VIF function 
library(scales) # for extra plot functions 
library(gridExtra) # extra plot functions (grid functions)
library(grid) # extra plot functions (grids)
```

## 

```{r}
# ---- loading data 
DDoS_dataset <- read_csv("Data/Friday-WorkingHours-Afternoon-DDos.pcap_ISCX.csv") %>% clean_names()

# view data 
head(DDoS_dataset)
```

## 3. Data Preprocessing & Preparation 

In this section, we focus primarily on preparing the dataset for analysis and modeling. Data processing is often considered the most crucial stage, as the quality of data directly impacts the accuracy of our model, reliability of the results, reducing overfitting, and mitigation of any unccessary errors. This process includes:

**Handling missing values:** Identifying and addressing any missing data to ensure a complete dataset, this also involves dropping columns/variables that contains any extreme amount of null values.

**Data transformation:** Converting variables to their respective formats, and possibly scaling the features.

**Handling class imbalance:** Addressing imbalanced variables often provides beneficial outcomes, as often if a column is imbalanced it may skew the results. In this case, handling this before the modeling phase is essential, in order to mitigate any future issues.

**Feature selection:** Removing any redundant or non-informative variables that do not contribute to the models performance and results. This includes unique identifying variables such as name, ID, etc.

**Removing highly correlated variables:** Identifying and removing variables that are highly correlated is essential to prevent redundancy, and multicollinearity, which can distort model interpretation and inflate our variance. This would also require us in utilizing VIF (Variance Inflation Factor) later on to track if our model overfitting, and impact the performance.

#### **3.1 Handling any possible missing values:** 

```{r}
# --- check for null values 
cat(paste("Total null values found:", sum(is.na(DDoS_dataset))))
```

#### **3.2 Data transformation:**

-   The label class, which is our target variable is currently labeled as DDoS, and BENIGN. We would then re-code as DDoS = 1, and BENIGN = 0.

-   Converting several variables as factors, as they represent categorical data that will be used in the modeling stage.

```{r}
# recode label variable to 1 and 0.
DDoS_dataset <- DDoS_dataset %>%
  mutate(label = dplyr::recode(label, "DDoS" = 1, "BENIGN" = 0))
cat("Recoded label variable to 1 and 0.\n")

# converting several variables to factor 
DDoS_dataset <- DDoS_dataset %>%
  mutate(
    label = as.factor(label),
    protocol = as.factor(protocol),
    fwd_psh_flags = as.factor(fwd_psh_flags),
    bwd_psh_flags = as.factor(bwd_psh_flags),
    fwd_urg_flags = as.factor(fwd_urg_flags),
    bwd_urg_flags = as.factor(bwd_urg_flags),
    fin_flag_count = as.factor(fin_flag_count),
    syn_flag_count = as.factor(syn_flag_count),
    rst_flag_count = as.factor(rst_flag_count),
    psh_flag_count = as.factor(psh_flag_count),
    urg_flag_count = as.factor(urg_flag_count),
    cwe_flag_count = as.factor(cwe_flag_count),
    ece_flag_count = as.factor(ece_flag_count),
  )
cat("Converted relevant variables to factors.\n")

```

#### **3.3 Handling class imbalance:** 

"As mentioned before, handling class imbalance is a vital process, often requiring an in-depth analysis of the underlying factors contributing to the imbalance. In the context of the DDoS dataset, one of the primary features where imbalance is observed is in the **protocol** column. This column likely contains values representing different network protocols (e.g., TCP, UDP, ICMP) used in network traffic.

A closer look at the distribution of values within the **protocol** column reveals significant disparities between different classes. For example, in one set of data, we might observe counts like the following:

In this case, the **BENIGN** class (label 0) is heavily underrepresented for certain protocol types, while **DDoS** attacks (label 1) dominate specific protocols. This imbalance is further amplified when considering the disproportionate distribution of attacks across different protocols. The **DDoS** class, for instance, may have an overwhelming presence in some protocols (such as Protocol 1), while others may show almost no attacks. This indicates that certain network protocols are more susceptible to DDoS attacks, which has critical implications for identifying and mitigating potential threats in real-world settings.

Thus, an effective approach to addressing this imbalance must consider the **protocol** column's distribution. Techniques like **oversampling**, **undersampling**, or **adjusting class weights** during model training could be employed to mitigate the impact of this imbalance. Additionally, **feature engineering** methods such as protocol-specific weighting could help the model focus more on the minority class (e.g., DDoS) while still capturing key information across all protocols."

In this, we can observe that **BENIGN class (label 0) is heavily underrepresented certain cases when the protocol used is "6". However, this is not the case for the DDoS (label 1), when the protocol used is "0" and "17",**

```{r}
# 
table(DDoS_dataset$label, DDoS_dataset$protocol)
```

```{r}
# bar plot of target and protocol distribution
ggplot(DDoS_dataset, aes(x = as.factor(protocol), fill = as.factor(label))) +
  geom_bar(color = "black", alpha = 0.7, linewidth = 0.3) +
  labs(title = "Figure 2: Distribution of DDoS Attack Labels by Protocol", x = "Protocol", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), 
                    labels = c("0" = "BENIGN", "1" = "DDoS"),
                    guide = guide_legend(title = NULL)) +
  scale_y_continuous(labels = scales::comma)

# Issues protocol variable 
# When "0", BENIGN is 54, while DDoS is 0. 
# When "6", BENIGN is 64793, while DDoS is 128027. 
# When "17", BENIGN is 32871, while DDoS is 0. 
```

```{r}
# Define the plots for syn_flag_count
plot1 <- ggplot(DDoS_dataset, aes(x = as.factor(syn_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "Syn Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for rst_flag_count
plot2 <- ggplot(DDoS_dataset, aes(x = as.factor(rst_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "RST Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for psh_flag_count
plot3 <- ggplot(DDoS_dataset, aes(x = as.factor(psh_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "PSH Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for ack_flag_count
plot4 <- ggplot(DDoS_dataset, aes(x = as.factor(ack_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "ACK Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for urg_flag_count
plot5 <- ggplot(DDoS_dataset, aes(x = as.factor(urg_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "URG Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for cwe_flag_count
plot6 <- ggplot(DDoS_dataset, aes(x = as.factor(cwe_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "CWE Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for ece_flag_count
plot7 <- ggplot(DDoS_dataset, aes(x = as.factor(ece_flag_count), fill = as.factor(label))) +    
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "ECE Flag Count", y = "Count", fill = "Label") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = label_comma()) +  
  theme(legend.position = "none", axis.title.x = element_text(size = 9))

# Define the plots for dummyplot1 to extract legend 
dummyplot1 <- ggplot(DDoS_dataset, aes(x = as.factor(syn_flag_count), fill = as.factor(label))) +
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"),  
                    labels = c("0" = "BENIGN", "1" = "DDoS"),
                    guide = guide_legend(title = NULL)) + 
  theme_minimal() + 
  guides(fill = guide_legend(title = "Label", 
                             ncol = 2, 
                             title.position = "top", 
                             title.hjust = 0.5))  
# function to extract the legend
g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  tmp$grobs[[leg]]
}
legend <- g_legend(dummyplot1)
```

```{r}
# Arrange the plots into a grid, leaving the last column for the legend
grid.arrange(
  plot1, plot2, plot3, plot4, plot5, plot6, plot7, legend,
  ncol = 2,  # Set 3 columns
  top = textGrob("Figure 3: Distribution of Flag Counts By Label", gp = gpar(fontsize = 10, fontface = "bold"))
)
```

```{r}
# Helper function to extract the legend
g_legend <- function(a.gplot) {
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  tmp$grobs[[leg]]
}

# Define the plots for fwd_psh_flags 
plot_1 <- ggplot(DDoS_dataset, aes(x = fwd_psh_flags, fill = as.factor(label))) +   
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "FWD PSH Flags", y = "Count", fill = "Label") +
  scale_fill_manual(values = c("blue", "red")) + 
  theme_minimal() +
  ggtitle("") +
  scale_y_continuous(labels = label_comma()) +
  theme(legend.position = "none")  # Remove the legend from this plot

# Define the plots for bwd_psh_flags 
plot_2 <- ggplot(DDoS_dataset, aes(x = bwd_psh_flags, fill = as.factor(label))) +   
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "BWD PSH Flags", y = "Count", fill = "Label") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  ggtitle("") +
  scale_y_continuous(labels = label_comma()) +
  theme(legend.position = "none")

# Define the plots for fwd_urg_flags 
plot_3 <- ggplot(DDoS_dataset, aes(x = fwd_urg_flags, fill = as.factor(label))) +   
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "FWD URG Flags", y = "Count", fill = "Label") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  ggtitle("") +
  scale_y_continuous(labels = label_comma()) +
  theme(legend.position = "none")

# Define the plots for bwd_urg_flags 
plot_4 <- ggplot(DDoS_dataset, aes(x = bwd_urg_flags, fill = as.factor(label))) +   
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  labs(x = "BWD URG Flags", y = "Count", fill = "Label") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  ggtitle("") +
  scale_y_continuous(labels = label_comma()) +
  theme(legend.position = "none")

# Define the plots for dummyplot2  
dummyplot2 <- ggplot(DDoS_dataset, aes(x = fwd_psh_flags, fill = as.factor(label))) +
  geom_bar(stat = "count", position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), 
                    labels = c("0" = "BENIGN", "1" = "DDoS"),
                    guide = guide_legend(title = NULL)) +
  theme_minimal() +
  guides(fill = guide_legend(title = "Label", ncol = 2))

# Extract the legend from the dummy plot
legend <- g_legend(dummyplot2)
```

```{r}
# Arrange the plots into a grid and add the title
grid.arrange(
  plot_1, plot_2, plot_3, plot_4,
  ncol = 2,
  top = textGrob("Figure 4: Distribution of PSH/URG Flags in the FWD & BWD", gp = gpar(fontsize = 10, fontface = "bold")),  # Adding a title at the top
  bottom = legend  # Adding the shared horizontal legend at the bottom
)
```

```{r}

  
# dropping due to imbalance issue 
' 14 total 
syn_flag_count
urg_flag_count
ece_flag_count
rst_flag_count
cwe_flag_count
fwd_psh_flags 
bwd_psh_flags
fwd_urg_flags
bwd_urg_flags
fwd_avg_bytes_bulk
fwd_avg_packets_bulk
fwd_avg_bulk_rate
bwd_avg_bytes_bulk
bwd_avg_bulk_rate' 


# check values for each of these variables 
summary(DDoS_dataset)

DDoS_dataset %>% 
  select(fwd_psh_flags, fwd_urg_flags, bwd_urg_flags, fwd_avg_bytes_bulk, 
         fwd_avg_packets_bulk, fwd_avg_bulk_rate, bwd_avg_bytes_bulk, bwd_avg_bulk_rate) %>%
  summary()

# we note that in this section we will explore a few variables that will be dropped due to their lack of importance in the later stage of analysis and modeling building. This will further include variables that contain only zero, and at later stage, we will conduct further analysis on any imbalanced variables. 

lm <- glm(label ~ flow_duration, data = DDoS_dataset, family = "binomial")

summary(lm)

unique(DDoS_dataset$label)
```

we note that in this section we will explore a few variables that will be dropped due to their lack of importance in the later stage of analysis and modeling building. This will further include variables that contain only zero, and at later stage, we will conduct further analysis on any imbalanced variables.

Remove unique ID columns

-   flow_id

-   source_ip

-   source_port

-   destination_ip

-   destination_port

-   timestamp, keeping for analysis purposes, will be dropped during model building.

```{r}
# --- remove unwanted variables 
DDoS_dataset <- DDoS_dataset %>%
  select(-flow_id, -source_ip, 
         -source_port, -destination_ip, 
         -destination_port)
```

## 4. Data Analysis

### 4.1 Correlation Matrix

In this section, creating a correlation matrix

### 4.2 Data Viz

```{r}
# Plot the distribution of the 'Target' variable

# bar plot of target distribution  
ggplot(DDoS_dataset, aes(x = as.factor(label))) +
  geom_bar(aes(fill = as.factor(label)), color = "black", alpha = 0.8, linewidth = 0.3) +
  labs(title = "Distribution of DDoS Attack Labels", x = "", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), 
                    labels = c("0" = "BENIGN", "1" = "DDoS"),
                    guide = guide_legend(title = NULL)) +  # Remove legend title
  scale_y_continuous(labels = scales::comma)


```

```{r}
ggplot() + 
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 0, ], 
             aes(x = flow_duration, y = fwd_iat_total, color = "BENIGN"), 
             alpha = 0.6, size = 2) +  # Color for label 0
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 1, ], 
             aes(x = flow_duration, y = fwd_iat_total, color = "DDoS"), 
             alpha = 0.8, size = 2.5) +  # Color for label 1
  scale_color_manual(values = c("BENIGN" = "blue", "DDoS" = "red")) +  # Custom color mapping
  theme_minimal() +
  labs(title = "Flow Duration vs FWD IAT Total", x = "Flow Duration", y = "FWD IAT Total") + 
  theme(legend.title = element_blank()) 

```

```{r}

ggplot() + 
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 0, ], 
             aes(x = total_fwd_packets, y = total_backward_packets, color = "BENIGN"), 
             alpha = 0.6, size = 2) +  # Color for label 0
  geom_point(data = DDoS_dataset[DDoS_dataset$label == 1, ], 
             aes(x = total_fwd_packets, y = total_backward_packets, color = "DDoS"), 
             alpha = 0.8, size = 2.5) +  # Color for label 1
  scale_color_manual(values = c("BENIGN" = "blue", "DDoS" = "red")) +  # Custom color mapping
  theme_minimal() +
  labs(title = "Total Forward Packets vs Total Backward Packets", 
       x = "Total Forward Packets", y = "Total Backward Packets") + 
  theme(legend.title = element_blank())  # Remove legend title

```

```{r}

ggplot(DDoS_dataset, aes(x = packet_length_std, y = packet_length_variance, colour = as.factor(label))) + 
  geom_point(alpha = 0.6, size = 2, position = "identity") +  # Set transparency for all points
  scale_color_manual(values = c("1" = "red", "0" = "blue")) +  # Reverse color mapping
  theme_minimal() +  # Clean theme
  labs(title = "Packet Length Std vs Packet Length Variance", x = "Packet Length Std", y = "Packet Length Variance Total") +  # Labels
  theme(legend.title = element_blank(),  # Remove legend title
        legend.key = element_rect(color = "black"))
```

```{r}




numeric_columns <- DDoS_dataset %>% select(where(is.numeric))

# Compute the correlation matrix for numeric variables
matriz_cor <- cor(numeric_columns)

sd_values <- sapply(numeric_columns, sd)

# Check which variables have a standard deviation of zero
zero_sd_vars <- names(sd_values[sd_values == 0])

# Print the variables with zero standard deviation
print(zero_sd_vars)

# fwd_psh_flags, 

# drop as only contains 0's and will not be important for the project 
# bwd_psh_flags, fwd_urg_flags, bwd_urg_flags



```

## 5. Reference list

-   Zayo Group (2024) *Average ddos attack cost businesses nearly half a million dollars in 2023, according to New Zayo Data: Press release: Zayo*, *Zayo.com*. Available at: <https://www.zayo.com/newsroom/average-ddos-attack-cost-businesses-nearly-half-a-million-dollars-in-2023-according-to-new-zayo-data/> (Accessed: 18 March 2025).

-   Garcia, D.M. (2021) *The 2011 PlayStation Network Hack – what actually happened?*, *WestSide Story*. Available at: <https://wsswired.com/4837/entertainment-3/the-2011-playstation-network-hack-what-actually-happened/> (Accessed: 18 March 2025).

## 6. Index 
