---
title: "DDOS Project EDA"
author: "Firass Elhouat"
date: "2025-03-14"
output: pdf_document
---

## 1. Introduction

In today’s digital landscape, Distributed Denial-of-Service (DDoS) attacks have emerged as one of the most prevalent and disruptive cybersecurity threats facing business and organizations of all sizes. This includes government agencies, educational institutions, healthcare, and any business that relies on servers and network access.

These attacks are designed to overwhelm a target’s/victim’s servers or network infrastructure with a flood of illegitimate traffic, rendering online services inaccessible to legitimate users and other stakeholders. The costs incurred during these attacks tend to vary depending on the duration and attackers motives. For instance, an insight report published in 2023 by Zayn Group, found that DDoS attacks on average last around 68 minutes, with unprotected organizations having to payout around \$6,000 per minute of each attack, [(Zayo Group (2024))](https://www.zayo.com/newsroom/average-ddos-attack-cost-businesses-nearly-half-a-million-dollars-in-2023-according-to-new-zayo-data/).

This totals to around \$408,000 per attack, that a business loses, which takes into account lost revenue, cost of detection, recovery, and cybersecurity preventives. While the immediate impact of a DDoS attack manifest as downtime or service disruptions, the broader consequences extend far beyond revenue loss. Businesses suffer from brand reputation damage, operational disruptions, legal ramifications, customer churn, and costly mitigation efforts.

A notable example is Sony’s PlayStation Network (PSN), which has been a frequent target of DDoS and other Malware attacks, most infamously in 2011, when a large-scale attack brought down the network for nearly a month, [(Garcia, D.M. (2021))](https://wsswired.com/4837/entertainment-3/the-2011-playstation-network-hack-what-actually-happened/) . This attack not only resulted in financial losses due to service interruptions and compensations to users but also severely impacted customer trust and brand loyalty. Many frustrated gamers switched to Steam (PC Gaming Platform) and Microsoft’s Xbox Live, which was perceived as more stable and secure at the time.

The PlayStation Network attack highlighted the long-term effects of DDoS attack, where the loss of consumer confidence led to a direct increase in competitors market share. Sony had to invest heavily in cybersecurity upgrades, public relations efforts, and damage control to regain trust, but the attack had already left a lasting impression on its user baser.

To better understand and model DDoS attacks, we will be utilizing a publicly available dataset from [Kaggle](https://www.kaggle.com/datasets/aymenabb/ddos-evaluation-dataset-cic-ddos2019/data){.uri}. This dataset contains extensive network traffic data, including packet details, communication protocols, and attack indicators. By applying Exploratory Data Analysis (EDA), we aim to identify patterns that distinguish normal traffic from malicious DDoS activity. Through this analysis, we aim to develop a predictive model that can detect potential DDoS threats in real time, helping us fully understand how these attacks occur, and what are some possible mitigation efforts that can be implemented to reduce the significant effects, and how each variable in our dataset explains our target variable.

## 2. Dataset

To better understand and model DDoS attacks, we will use the CIC-DDoS2019 dataset contains 225,745 rows, and 85 various variables that help differentiate between normal and attack-related traffic. The dataset includes attributes such as:

-   Identifier and date variables: These are most likely to be dropped as they are not useful during the modeling phase of our project.

    -   flow_id

    -   source_ip

    -   source_port

    -   destination_ip

    -   destination_port

    -   Time stamp

-   Categorical Variables:

    -   **protocol:**

-   Continuous Variables:

    -   **flow_duration:**

    -   **total_fwd_packets:**

    -   **total_backward_packets:**

    -   **total_length_of_fwd_packets:**

    -   **total_length_of_bwd_packets:**

    -   FWD PACKET LENGTH:

        -   **fwd_packet_length_max:**

        -   **fwd_packet_length_min:**

        -   **fwd_packet_length_mean:**

        -   **fwd_packet_length_std:**

    -   BWD PACKET LENGTH:

        -   **bwd_packet_length_max:**

        -   **bwd_packet_length_min:**

        -   **bwd_packet_length_mean:**

        -   **bwd_packet_length_std:**

    -   **flow_bytes_s:**

    -   **flow_packets_s:**

    -   Flow IAT:

        -   **flow_iat_mean:**

        -   **flow_iat_std:**

        -   **flow_iat_max:**

        -   **flow_iat_min:**

    -   FWD IAT:

        -   **fwd_iat_total:**

        -   **fwd_iat_mean:**

        -   **fwd_iat_std:**

        -   **fwd_iat_max:**

        -   **fwd_iat_min:**

    -   BWD IAT:

        -   **bwd_iat_total:**

        -   **bwd_iat_mean:**

        -   **bwd_iat_std:**

        -   **bwd_iat_max:**

        -   **bwd_iat_min:**

    -   **fwd_psh_flags:**

    -   **bwd_psh_flags:**

    -   **fwd_urg_flags:**

    -   bwd_urg_flags:

    -   **fwd_header_length_41:**

    -   **bwd_header_length:**

    -   **fwd_packets_s:**

    -   **bwd_packets_s:**

    -   PACKET LENGTH

        -   **min_packet_length:**

        -   **max_packet_length:**

        -   **packet_length_mean:**

        -   **packet_length_std:**

        -   **packet_length_variance:**

    -   **fin_flag_count:**

    -   **syn_flag_count:**

    -   **rst_flag_count:**

    -   **psh_flag_count:**

    -   **ack_flag_count:**

    -   **urg_flag_count:**

    -   **cwe_flag_count:**

    -   **ece_flag_count:**

    -   **down_up_ratio:**

    -   **average_packet_size:**

    -   **avg_fwd_segment_size:**

    -   **avg_bwd_segment_size:**

    -   **fwd_header_length_62:**

    -   **fwd_avg_bytes_bulk:**

    -   **fwd_avg_packets_bulk:**

    -   **fwd_avg_bulk_rate:**

    -   **bwd_avg_bytes_bulk:**

    -   **bwd_avg_packets_bulk:**

    -   **bwd_avg_bulk_rate:**

    -   **subflow_fwd_packets:**

    -   **subflow_fwd_bytes:**

    -   **subflow_bwd_packets:**

    -   **subflow_bwd_bytes:**

    -   **init_win_bytes_forward:**

    -   **init_win_bytes_backward:**

    -   **act_data_pkt_fwd:**

    -   **min_seg_size_forward:**

    -   ACTIVE:

        -   **active_mean:**

        -   **active_std:**

        -   **active_max:**

        -   **active_min:**

    ```{=html}
    <!-- -->
    ```
    -   IDLE:

        -   **idle_mean:**

        -   **idle_std:**

        -   **idle_max:**

        -   **idle_min:**

    -   

-   Response Variable:

    -    label:

```{r setup, include=FALSE}
# --- loading libraries 
library(tidyverse) # for extra functions 
library(janitor) # for column name cleaning 
library(viridis) # for color palettes

```

## 

```{r}
# ---- loading data 
DDoS_dataset <- read_csv("Data/Friday-WorkingHours-Afternoon-DDos.pcap_ISCX.csv") %>% clean_names()

# view data 
print(DDoS_dataset)
```

```{r}
# unique Values
length(unique(DDoS_dataset$timestamp))
length(unique(DDoS_dataset$fwd_header_length))

# remove unique ID columns 
# (flow_id, source_ip, source_port, destination_ip, destination_port)
```

## 3. Data Preprocessing 

Remove unique ID columns

-   flow_id

-   source_ip

-   source_port

-   destination_ip

-   destination_port

-   timestamp

```{r}
# --- remove unwanted variables 

```

Checking and removing null rows

```{r}
# --- check for null values 

```

## 4. Data Analysis 

### 4.1 Correlation Matrix 

In this section, creating a correlation matrix

```{r}

```

### 4.2 Data Viz 

```{r}
# Plot the distribution of the 'Target' variable

# bar plot of target distribution  
ggplot(DDoS_dataset, aes(x = as.factor(label))) +
  geom_bar(aes(fill = as.factor(label)), color = "black") +
  labs(title = "Distribution of DDoS Attack Labels", x = "", y = "Count") +
  theme_minimal() +
  scale_fill_viridis(discrete = TRUE,option = "cividis", direction = -1) +  # Using the Viridis color palette
  scale_y_continuous(labels = scales::comma)

table(DDoS_dataset$label, DDoS_dataset$protocol)


# bar plot of target and protocol distribution
ggplot(DDoS_dataset, aes(x = as.factor(protocol), fill = as.factor(label))) +
  geom_bar(color = "black") +
  labs(title = "Distribution of DDoS Attack Labels by Protocol", x = "", y = "Count") +
  theme_minimal() +
  scale_fill_viridis(discrete = TRUE, option = "cividis", direction = -1) +
  scale_y_continuous(labels = scales::comma)

# Issues protocol variable 
# When "0", BENIGN is 54, while DDoS is 0. 
# When "6", BENIGN is 64793, while DDoS is 128027. 
# When "17", BENIGN is 32871, while DDoS is 0. 

# Highly correleted with BENIGN, especially when the value is 0 or 17. 
```

## 5. Reference list 

-   Zayo Group (2024) *Average ddos attack cost businesses nearly half a million dollars in 2023, according to New Zayo Data: Press release: Zayo*, *Zayo.com*. Available at: <https://www.zayo.com/newsroom/average-ddos-attack-cost-businesses-nearly-half-a-million-dollars-in-2023-according-to-new-zayo-data/> (Accessed: 18 March 2025).

-   Garcia, D.M. (2021) *The 2011 PlayStation Network Hack – what actually happened?*, *WestSide Story*. Available at: <https://wsswired.com/4837/entertainment-3/the-2011-playstation-network-hack-what-actually-happened/> (Accessed: 18 March 2025).
